---
title: "Untangling messy records with `set_group_id()`"
subtitle: "Same person, three different rows — graphfast finds them in one call"
---

```{r}
#| label: setup
#| message: false
#| warning: false
library(graphfast)
library(data.table)
set.seed(1)
```

::: {.callout-note}
This packages uses the `graphfast` package. [Visit repo →](https://github.com/KyleHaynes/graphfast)
:::

## The problem

Real-world data is full of duplicates that don't *look* like duplicates. A
customer signs up three times with three different names, but the same phone
number creeps into two of those records and the same email into the other two.
No single column ties all three rows together — but a **chain** of shared
values does.

This is exactly the kind of problem graph algorithms are good at, and it's why
`set_group_id()` exists: treat every row and every value it contains as nodes
in a graph, connect a row to each value it holds, and then ask "which nodes
end up in the same connected component?" Rows in the same component get the
same group id.

## Why it's quick to use

No graph code, no manual joins — just point it at the columns that might
overlap:

```{r}
dt <- data.table(
  id     = 1:5,
  name   = c("John Smith", "J. Smith", "Jane Doe", "Bob Jones", "Robert Jones"),
  phone  = c("555-0101", "555-0101", "555-0202", "555-0303", "555-0404"),
  email  = c("john@mail.com", "jsmith@mail.com", "jane@mail.com",
             "bob@mail.com", "555-0303")
)
dt
```

Rows 1 and 2 share a phone number. Row 5 doesn't share a phone or name with
row 4 at all — but its email happens to be the same string as row 4's phone
number. One call resolves the whole chain:

```{r}
set_group_id(dt, cols = "phone|email", var_output_name = "gid")
```

Rows 1 and 2 collapse into one group (matching phone numbers), and rows 4 and
5 collapse into another (row 5's stray value matches row 4's phone), while
row 3 (Jane Doe) stays on her own. That's the whole API: a data.table, a
column pattern, done.

## Isolation: stopping false links across columns

Notice what just happened with row 5 — its email value (`"555-0303"`) is
*textually identical* to row 4's phone number, so they got linked even though
one is "obviously" a phone number and the other is "obviously" an email. That
might be exactly what you want (data entry errors do happen), or it might be a
false positive you need to prevent.

That's what `isolate` is for: it keeps values from different column sets in
separate namespaces, so identical text only links rows **within the same
set**.

```{r}
dt2 <- copy(dt)[, gid := NULL]
set_group_id(dt2, isolate = list("phone", "email"))
```

With `phone` and `email` isolated from each other, row 5's stray value no
longer connects it to row 4 — the phone-shaped string sitting in the email
column is just ignored for grouping purposes. Row 1 and 2 still merge,
because that link is a genuine phone match within the same set.

This matters most when you're matching across many heterogeneous columns
(`phone1`, `phone2`, `alt_phone`, `email`, `secondary_email`, ...) and don't
want a coincidental string collision between unrelated fields to merge two
otherwise-unrelated customers.

## Seeing the linkage

Because `set_group_id()` can hand back the record-to-record edge list (every
pair of ids that share a value, plus the shared value itself), the result
drops straight into [`visNetwork`](https://datastorm-open.github.io/visNetwork/)
for a quick visual sanity check:

```{r}
#| eval: !expr requireNamespace("visNetwork", quietly = TRUE)
library(visNetwork)

dt3 <- copy(dt)[, gid := NULL]
edges <- set_group_id(dt3, isolate = list("phone", "email"), return_edges = TRUE)

nodes <- data.frame(
  id    = dt3$id,
  label = dt3$name,
  group = dt3$gid,
  title = paste0("phone: ", dt3$phone, "<br>email: ", dt3$email)
)
vis_edges <- data.frame(
  from  = edges$from,
  to    = edges$to,
  title = edges$value   # hover shows the value that links the pair
)

visNetwork(nodes, vis_edges, width = "100%", height = "320px") |>
  visNodes(shape = "dot", size = 24) |>
  visOptions(highlightNearest = TRUE) |>
  visLegend()
```

Two clusters, one isolated node — exactly what the table above said, but now
it's obvious at a glance which shared value pulled which records together
(hover an edge to see it).

## Takeaways

- **Quick**: one function call, no manual graph wrangling, works directly on
  a `data.table` by reference.
- **Flexible matching**: a regex (`cols`) picks up every relevant column at
  once, so adding a new identifier column doesn't require new code.
- **Safe by default, configurable when needed**: `isolate` stops unrelated
  columns from cross-linking on a coincidental string match, while still
  letting genuinely shared values (within a set) merge records.
- **Inspectable**: `return_edges = TRUE` gives you the *why* behind every
  group, which is what makes the `visNetwork` plot above possible.

See `?set_group_id` and the [examples page](docs/examples.qmd) for more
patterns, including `incomparables` for excluding sentinel values like
`"Unknown"` from grouping.
