---
title: "Three string functions in `fast.string` that earn their keep"
subtitle: "Drop-in replacements for base R string ops, parallelised in C++ — sometimes 100×+ faster"
author: "Kyle Haynes"
date: "2026-06-28"
format:
  html:
    toc: true
    code-fold: false
    df-print: kable
execute:
  eval: false
---


::: {.callout-note}
This packages uses the `fast.string` package. [Visit repo →](https://github.com/KyleHaynes/faststring)
:::

Base R's string functions are written for correctness and generality, not raw
speed. On a 10-row vector that's invisible. On a few million rows — the kind of
data that turns up in record linkage, log parsing, or ETL — the wait gets real.

`fast.string` re-implements a handful of the hottest string operations in C++
(PCRE2 for regex, RE2's zero-copy `StringPiece` for fixed strings) and runs them
in parallel across every core via Intel TBB. The functions are **drop-in
replacements** — same arguments, same results — so adopting them is usually a
one-character change.

Here are three that pull the most weight, with demos and benchmarks you can run
yourself.

```{r setup}
library(fast.string)
library(microbenchmark)
```

> All timings below are medians over 10 runs on a 16-core machine, R 4.5.0.
> Your mileage will vary with core count, but the shape holds.

---

## 1. `fgrepl()` — pattern matching without the wait

`fgrepl()` is `grepl()`. Same `pattern`, `x`, `ignore.case`, `fixed`, `perl`
arguments; same logical vector out. The only difference is it spreads the work
across your cores. If you hand it a pattern that uses PCRE-only syntax
(lookaheads, backreferences, …) it quietly delegates to base R so you never get
a wrong answer.

```{r fgrepl-demo}
x <- sample(c("apple pie", "banana bread", "cherry cake"), 2e6, replace = TRUE)

# The only change from base R: prepend "f"
fgrepl("berry", x)        # vs grepl("berry", x)

identical(grepl("berry", x), fgrepl("berry", x))
#> [1] TRUE
```

**Benchmark** (2,000,000 strings):

```{r fgrepl-bench}
microbenchmark(
  base        = grepl("berry", x),
  fast.string = fgrepl("berry", x),
  times = 10
)
```

| function | median |
|----------|-------:|
| `grepl()` | 339 ms |
| `fgrepl()` | **16 ms** |

**~21× faster**, bit-for-bit identical output. The win comes for free — no
change to how you write the call.

---

## 2. `format_date()` — the one with the absurd benchmark

Formatting dates in base R goes through locale handling, timezone lookups, and
`strptime` machinery you almost never need when you just want `"2024-03-17"`.
`format_date()` skips all of it and uses an integer date-math algorithm, with no
calendar object in sight.

It takes a `Date` (or raw days-since-epoch) and gives you one of four common
layouts: `"iso"`, `"compact"` (`YYYYMMDD`), `"dmy"`, or `"ymd_slash"`.

```{r date-demo}
d <- as.Date("2000-01-01") + sample(0:9000, 2e6, replace = TRUE)

format_date(d, "iso")      # "2003-11-24" ...
format_date(d, "compact")  # "20031124"   ...

identical(format(d, "%Y-%m-%d"), format_date(d, "iso"))
#> [1] TRUE
```

**Benchmark** (2,000,000 dates):

```{r date-bench}
microbenchmark(
  base        = format(d, "%Y-%m-%d"),
  fast.string = format_date(d, "iso"),
  times = 10
)
```

| function | median |
|----------|-------:|
| `format(d, "%Y-%m-%d")` | 14,953 ms |
| `format_date(d, "iso")` | **102 ms** |

**~147× faster.** Base R takes 15 *seconds*; `format_date()` takes a tenth of a
second. If you've ever watched a pipeline stall on a `format()` call over a big
date column, this is the fix. (There's a matching `fas.Date()` for the reverse
direction.)

---

## 3. `jaro_winkler()` / `jaro_winkler_tokens()` — fuzzy matching for record linkage

Comparing names and addresses across datasets means tolerating typos, so you
need a string-similarity score. `jaro_winkler()` computes the classic metric,
vectorised over pairs and parallelised:

```{r jw-demo}
jaro_winkler("haynes", "haines")
#> ~0.94
```

It matches `stringdist`'s implementation to floating-point precision while
running a few times faster — but the real reason to reach for this package is
`jaro_winkler_tokens()`, which has no base-R or `stringdist` equivalent.

Plain Jaro-Winkler reads each string as one ordered run of characters, so it
falls apart the moment words **move** or **split**, which is exactly what messy
name data does:

```{r jw-tokens-demo}
a <- c("Kyle John Haynes", "O'Brien",  "John Smith")
b <- c("John Kylie Haynes","O Brien",  "John Smith Jones")

data.frame(
  a, b,
  plain    = round(jaro_winkler(a, b), 3),
  tokens   = round(jaro_winkler_tokens(a, b), 3)
)
```

| a | b | plain | tokens |
|---|---|------:|-------:|
| Kyle John Haynes | John Kylie Haynes | 0.897 | **0.984** |
| O'Brien | O Brien | 0.914 | **1.000** |
| John Smith | John Smith Jones | 0.925 | 0.929 |

`jaro_winkler_tokens()` splits on whitespace, scores every token against every
token, greedily pairs the best matches, and also tries a punctuation-stripped
collapsed comparison — keeping the higher score. Reordered names
(`"Kyle John"` ↔ `"John Kylie"`) and stray punctuation (`"O'Brien"` ↔
`"O Brien"`) stop being penalised, while a genuinely extra token (`"Jones"`)
still costs a little. It'll even rescue fused words like `"KYLEJOHN"` ↔
`"KYLE JOHN"` with `contractions = TRUE`.

**Benchmark** (500,000 pairs, vs `stringdist`):

```{r jw-bench}
library(stringdist)
microbenchmark(
  stringdist  = 1 - stringdist(a, b, method = "jw", p = 0.1),
  fast.string = jaro_winkler(a, b, p = 0.1),
  times = 10
)
```

| function | median |
|----------|-------:|
| `stringdist(method="jw")` | 15.0 ms |
| `jaro_winkler()` | **5.3 ms** |

**~2.8× faster than an already-compiled C implementation**, with a max
difference of `1e-16` — and the token-aware variant gives you matching power the
alternatives simply don't have.

---

## The pattern

| base R | `fast.string` | speedup\* |
|--------|---------------|----------:|
| `grepl()` | `fgrepl()` | ~21× |
| `format(d, "%Y-%m-%d")` | `format_date()` | ~147× |
| `stringdist(method="jw")` | `jaro_winkler()` | ~2.8× |

\* 16 cores, large vectors; see benchmarks above.

These are three of several — the package also has `gsub_all()` (multi-pattern
substitution in one pass), parallel `ftrimws()`/`fsubstr()`/`fnchar()`/
`fchartr()`, `date_parts()`, and phonetic codings (`soundex()`, `nysiis()`).

The common thread: **same arguments, same results, a fraction of the time.**
For interactive work on small data it won't matter. For production pipelines
over millions of rows, swapping in a few of these can turn a coffee break back
into a keystroke.
