---
title: "Kyle Haynes: Curriculum Vitae"
format:
  html:
    toc: true
  pdf:
    author: Kyle Haynes
    pdf-engine: weasyprint
css: cv.css
format-links: [pdf]
---

::: {.print-only .contact-block}
kyle@kylehaynes.com.au
:::

```{r}
# require(weasydoc)
```

## EDUCATION

**Doctor of Philosophy -- Econometrics and Business Statistics** <br> *Monash University* [*Aug 2021 --- Present*]{.cvdate}

-   New principles and methods for complex data preparation and integration, with applications to official statistics, web-scrapped data and satellite raster images
-   Monash Data Futures Institute PhD Top-Up Scholarship (2021-2024)
-   Expected submission date: Nov 2024

## WORK EXPERIENCE{style="page-break-before: always"}

### Senior Statistician, A06

***Queensland Government Statisticians Office (QGSO): Data Integration*** [Jan 2020 -- Present]{.cvdate}

Update Update Update.

-   Update


### Statistician, A05

***QGSO: Statistical Collection Intergarion & Analysis*** [Jan 2020 -- Present]{.cvdate}

Update Update.

### Statistician, A04

***QGSO: Statistical Collection Intergarion & Analysis*** [Jan 2020 -- Present]{.cvdate}

Update Update.


::: {.no-print}
## PRESENTATIONS AND WORKSHOPS

-   Talk: [Misadventures with Reproducibility in R](https://github.com/cynthiahqy/rladies-2022-11) (30 Nov 2022, R Ladies Melbourne Meetup)
-   Talk: [Designing R Packages](https://github.com/cynthiahqy/design-r-magic_2022-10) (4 Oct 2022, Monash EBS Data Science Research Software Study Group)
-   Talk: [Quarto Websites as Research Compendiums](https://github.com/cynthiahqy/quarto-wiki-template) (16 Aug 2022, Monash EBS Data Science Research Software Study Group)
-   Workshop: [Writing academic papers with Rmarkdown and friends](https://www.cynthiahqy.com/posts/r-markdown-and-friends/) (9 Aug 2022, Monash University)
:::

::: {.print-only}
## References
Available upon request
:::


```{r}
head(iris)
require(data.table)


d <- data.table(iris)

head(d)



```

asdas 

<details>
asdas
</details>